Left-linear rules in adhesive categories

### 2024-02-09 icalp/fusione.tex
general pass

% full or conference version
\newif\iffull
\fulltrue
%\fullfalse

% additional text
\iffull
\newcommand{\full}[1]{#1}
%\newcommand{\short}[1]{}
\else 
\newcommand{\full}[1]{}
%\newcommand{\short}[1]{#1}
\fi

@@

\begin{abstract}
When two sequences of steps executed by a computational device can be considered equivalent?
This is a relevant question if one considers concurrent and distributed systems, since the independence
of two steps means that they could be executed on any order, and potentially in parallel.
%
We investigate the issue in the context of adhesive categories, a very
general instance of the DPO approach to rewriting. More precisely, we
focus on left-linear rules, which not only consume, preserve and generate entities,
as linear rules do, but may ``merge'' parts of the state.
%
An apparently minimal, yet technically hard enhancement, which allows to properly extend the 
formalisms that can be modelled in the adhesive framework, and for which we prove a 
Church-Rosser property.
%
% We refine the established notion of sequential independence between steps, and ...
 \end{abstract}

@@

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

@@

In particular, we can apply this result to slices over and under a given object.

\begin{corollary}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{corollary}

@@

Often, we are interested in derivations only up to some notion of coherent isomorphism between them. This is done with the help of \Cref{rem:func}.

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewritings. We
observe that the shift to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

As already mentioned, when viewing derivations as concurrent computations, 
for linear graph rewriting systems it is canonical to identify  derivations which are equal 
``up to switching''. The same notion can be given for left-linear rewriting systems relying on notion of switch.

\begin{definition}[switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \leftrightsquigarrow_{(i,i+1)} \der{E}$. Then \emph{switch
    equivalence} is the least equivalence on derivation such that if
  $\der{D} \leftrightsquigarrow_{(i,i+1)} \der{E}$ then
  $\der{D} \equiv^s \der{E}$.
\end{definition}

@@

The fact that sequential indepedence implies switchability always
holds for linear rules. [REF] The result is so fundamental that it is
not even stated explicitly, in the sense that switchability is not introduced
axiomatically as in Definition~\ref{def:switch}, but it is based on
the explicit construction of a switch.

@@

Switches along the same independence pair are unique up to abstraction
equivalence (see Proposition~\ref{thm:switch_uni}).

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@/^.20cm/@{.>}[ddrr]|(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@/_.20cm/@{.>}[ddll]|{u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1 &&
      \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@{>.>}[ull]^{p_0} \ar@{.>}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{1,2\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ (dotted part of the diagram).
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first and second square below are pushouts and if the third
  square exists and it is a pushout too
  \[
    \xymatrix{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}&K_1
      \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0} \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0&P \ar@{.>}[r]_{q_1} &
      Q_1}
  \]
\end{definition}

>>>>

% full or conference version
\newif\iffull
\fulltrue
\fullfalse

% additional text
\iffull
\newcommand{\full}[1]{{\color{blue}#1 full}}
%\newcommand{\short}[1]{}
\else 
\newcommand{\full}[1]{}
%\newcommand{\short}[1]{#1}
\fi

@@

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed on any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, which
  allows to properly extend the formalisms that can be modelled in the
  adhesive framework, and for which we prove a Church-Rosser property.
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  The canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to suitable well-behaved classes of left-linear
  rewriting systems.  Additionally, a fundamental characterisation of
  the notion of independence fails when moving to left-linear
  rewriting systems. This requires a substantial revision of the
  theory as in the linear case, a number of basic properties,
  essential in the development of the theory of concurrency, were
  derived from such a characterisation.
 \end{abstract}

@@

Here we observe that, in general, both properties can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

@@

\begin{theorem}\label{cor:slice}
  Let $X$ be an object of an $\mathcal{M}$-adhesive category
  $\X$. Then $\X/X$ and $X/\X$ are, respectively, $\mathcal{M}/X$-
  and $X/\mathcal{M}$-adhesive, where
  \[\mathcal{M}/X:=\{m\in \mathcal{A}(\X/X) \mid m\in \mathcal{M} \}
    \qquad X/\mathcal{M}:=\{m\in \mathcal{A}(X/\X) \mid m\in
    \mathcal{M} \}
  \]
\end{theorem}

@@

Often, we are interested in derivations only up to some notion of coherent isomorphism between them. This leads us to the following definition.

\noindent
\parbox{8cm}{
  \begin{definition}[abstraction equivalence]
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \Deltamin(\der{D})}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$are $\der{D}'$ are \emph{abstraction
      equivalent} if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R\ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{H_i}}}$}

@@

In this section we discuss the notion of independence between
consecutive rewriting steps, which is a fundamental ingredient of the
rewriting theory, which comes into play when viewing a derivation
sequence as a concurrent computation, the idea being that of
abstracting out the order of execution of independent rewritings. We
observe that moving to left-linear rules leads to the failure of
various basic properties of the classical notion of independence and
we single out a framework in which these can be re-established.

@@

It can be shown that switches along the same independence pair are unique up to abstraction
equivalence (see Proposition~\ref{thm:switch_uni}).

As already mentioned, for linear rewriting systems, when viewing
derivations as concurrent computations, it is canonical to identify
derivations which are equal ``up to switching''. The same notion can
be given for left-linear rewriting systems relying on notion of
switch.

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ we write
  $\der{D} \leftrightsquigarrow_{(i,i+1)} \der{E}$. Then \emph{switch
    equivalence} is the least equivalence on derivation such that if
  $\der{D} \leftrightsquigarrow_{(i,i+1)} \der{E}$ then
  $\der{D} \equiv^s \der{E}$.
\end{definition}

In words, we let $\der{D} \equiv^s \der{E}$ when one derivation can be obtained from the other via a sequence of switches.

@@

The fact that sequential indepedence implies switchability always
holds for linear rules (see \Cref{prop:equi}). The result is so fundamental that it is
not even stated, in the sense that switchability is not introduced
axiomatically as in Definition~\ref{def:switch}, but it is based on
the explicit construction of a switch.

@@

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let with $\X$ an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@/^.20cm/[ddrr]|(.3)\hole|(.55)\hole^(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@/_.20cm/[ddll]|(.3)\hole|(.55)\hole_(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1 &&
      \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>->}[ull]^{p_0} \ar@/_.3cm/[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first and second square below are pushouts and if the third
  square exists and it is a pushout too
  \[
    \xymatrix{
      %
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}&K_1
      \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0} \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0&P \ar@{.>}[r]_{q_1} &
      Q_1}
  \]
\end{definition}
